---
unit_title: "Spatial visualization"
theme: visualize
needs: [spatial-basics]
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup, message = FALSE}
library(tidyverse)
library(sf)
```

In _Spatial basics_, you learned how to plot spatial data using the `plot()` function. Now, we'll show you how to visualize spatial data using `geom_sf()`. `geom_sf()` has the advantage of being part of ggplot2, meaning you can layer geoms, use scales functions to change colors, easily tweak legends, alter the theme, etc.

## `geom_sf()`

The [ussf package](https://github.com/dcl-docs/ussf) makes it easy to create maps of the U.S. by supplying state, county, or commuting zone boundaries. Use `geography = "state"` to return boundaries for all U.S. states.

```{r}
states <- ussf::boundaries(geography = "state")

states %>% 
  select(GEOID, NAME, geometry)
```

We can use `geom_sf()` to plot the state boundaries.

```{r}
states %>% 
  ggplot() +
  geom_sf()
```

We don't have to supply any aesthetic mappings to `ggplot()`. `geom_sf()` will, by default, use the column that stores simple features data. This column will be of class `sfc` and typically be called `geometry`.

```{r}
class(states$geometry)
```

## Projections

The Earth is not flat, but your plots are. To create effective maps, you'll need to project a three-dimensional shape (the Earth) onto a two-dimensional surface (your screen). 

Latitude and longitude coordinates are unprojected. Each point picked out by a longitude and latitude combination specifies a point on an ellipsoid. This means that plots of large geographic areas, like the U.S., will look strange when plotted in longitude-latitude. 

```{r}
states_longlat <- 
  ussf::boundaries(geography = "state", projection = "longlat")
 
states_longlat %>% 
  filter(!(NAME %in% c("Alaska", "Hawaii"))) %>% 
  ggplot() +
  geom_sf()
```

A different projection will improve our map substantially. When mapping the entire U.S., we recommend U.S. Albers. U.S. Albers is the default for the ussf package. 

```{r}
states_albers <- ussf::boundaries(geography = "state")

states_albers %>% 
  ggplot() +
  geom_sf()
```

Anytime you project a three-dimensional object onto a two-dimensional object, some aspect of the object will be distorted. U.S. Albers chooses to accurately reflect area, at the cost of minimally distorting scale and shape. 

ussf also scales Alaska and Hawaii and places them below the continental U.S. This is not related to the Albers projection.

If you're visualizing a small geographic area, like North Carolina, you can just use longitude and latitude. The Earth is approximately flat for a small enough area.

```{r}
states_longlat %>% 
  filter(NAME == "North Carolina") %>% 
  ggplot() +
  geom_sf()
```

However, California covers a bit too much area for this to work. On its own, California also looks strange if projected with U.S. Albers. 

```{r}
states_albers %>% 
  filter(NAME == "California") %>%
  ggplot() +
  geom_sf()
```

Luckily, there's an Albers equal area projection just for California, called California Albers. You can specify the projection and coordinate reference system for a map using _PROJ strings_. Here's the PROJ string for California Albers.

```{r}
CA_ALBERS <- 
  "+proj=aea +lat_1=34 +lat_2=40.5 +lat_0=0 +lon_0=-120 +x_0=0 +y_0=-4000000 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
```

_aea_ stands for _Albers equal area_, and we are using the [WGS84 coordinate reference system](https://en.wikipedia.org/wiki/World_Geodetic_System). The [Spatial Reference](https://spatialreference.org/) website is a good place to look up PROJ strings. 

We can change the projection before plotting with `st_transform()`.

```{r}
states_albers %>% 
  filter(NAME == "California") %>% 
  st_transform(crs = CA_ALBERS) %>% 
  ggplot() +
  geom_sf() 
```

## Choropleths

Say we want to visualize population density across California counties. `population` contains population data for California counties. 

```{r include=FALSE}
query <- "https://api.census.gov/data/2018/pep/population?get=GEONAME,POP,DATE_DESC&for=county:*&in=state:06"
date <- "7/1/2018"

population <-
  query %>% 
  jsonlite::fromJSON() %>% 
  as_tibble() %>% 
  janitor::row_to_names(row_number = 1) %>% 
  filter(str_detect(DATE_DESC, date)) %>% 
  transmute(
    fips = str_c(state, county) %>% as.integer(),
    name = GEONAME,
    population = as.double(POP)
  ) %>% 
  arrange(fips)
```

```{r}
population
```

(We obtained the population data from the Census Bureau's Population Estimates API. See our chapter on [Census Bureau APIs](http://dcl-wrangle.stanford.edu/census.html) for more information.)

We'll join our population data with boundaries from `ussf::boundaries()`. Because we're looking at a smaller area, we can use a higher resolution than the default.

```{r}
FIPS_CA <- 6L

ca_counties <-
  ussf::boundaries(geography = "county", resolution = "5m") %>% 
  filter(as.integer(STATEFP) == FIPS_CA) %>% 
  st_transform(crs = CA_ALBERS) %>% 
  transmute(
    fips = as.integer(GEOID),
    area_land = ALAND
  ) %>% 
  left_join(population, by = "fips") %>% 
  select(fips, name, population, area_land) 

ca_counties
```

Note that you have to supply the boundaries data as the first argument of `left_join()`. If you supply it as the second, the tibble will lose its `sfc` class, and `geom_sf()` will no longer work as expected. 

Just like with other geoms, you can supply additional aesthetics to `geom_sf()`. For polygons like the counties of California, `fill` is the most useful aesthetic. Let's visualize population density by county.

`area_land` is in square meters, but we'll express density in terms of population per square mile.

```{r}
ca_counties <-
  ca_counties %>% 
  mutate(density = population / (area_land * 3.861e-7))
  
ca_counties %>% 
  ggplot(aes(fill = density)) +
  geom_sf(color = "white", size = 0.2)
```

Maps like this one, in which geographic areas are colored according to some variable, are called _choropleths_. 

Our map has a substantial problem. It's pretty much all the same color! It's very difficult to tell the difference in population density between counties, particularly for low-density counties. 

Let's look at the distribution of `density`.

```{r}
ca_counties %>% 
  ggplot(aes(density)) +
  geom_histogram(binwidth = 200) 
```

Most of the counties are low density, but there are some outliers. By default, our sequential color scale will linearly map colors between the minimum (`r format(min(ca_counties$density, na.rm = TRUE), digits = 3)`) and the maximum (`r format(max(ca_counties$density, na.rm = TRUE), digits = 3, big.mark = ",")`). (The extreme outlier is San Francisco County.)

One way to deal with this problem is to cap `density` using `pmin()`. Now, everything above our cutoff will be represented with the same color. 

```{r}
ca_counties %>% 
  mutate(density = pmin(density, 700)) %>% 
  ggplot(aes(fill = density)) +
  geom_sf() 
```

This map makes it much easier to see the high density areas and compare the lower density areas to each other. A better color scale and thinner borders will also improve our plot.

```{r}
ca_counties %>% 
  mutate(density = pmin(density, 700)) %>% 
  ggplot(aes(fill = density)) +
  geom_sf(size = 0.3) +
  scale_fill_gradientn(
    colors = RColorBrewer::brewer.pal(n = 9, name = "PuRd")
  )
```

You can find this lighter color scale, and many others, at the [ColorBrewer website](http://colorbrewer2.org/#). You can also browse all the palettes included in the RColorBrewer package by running `RColorBrewer::display.brewer.all()`.

If you do truncate a variable to improve your color scale, make sure to clearly indicate that you did so in your legend.

```{r}
labels <- function(x) {
  if_else(x < 700, as.character(x), "700+")
}

ca_counties %>% 
  mutate(density = pmin(density, 700)) %>% 
  ggplot(aes(fill = density)) +
  geom_sf(size = 0.3) +
  scale_fill_gradientn(
    breaks = scales::breaks_width(width = 200, offset = -100),
    labels = labels,
    colors = RColorBrewer::brewer.pal(n = 9, name = "PuRd")
  )  +
  theme_void() +
  labs(
    fill = "Population density\n(population / square mile)"
  )
```

`theme_void()` is an easy way to remove the grid lines and tick mark labels, which aren't necessary for someone to understand our map.

## Layering

Just like with other geoms, you can layer `geom_sf()` with additional `geom_sf()`s or other geoms.  

Previously, we've been plotting multipolygons with `geom_sf()`.

```{r}
class(ca_counties$geometry)
```

Recall that _sf_ stands for _simple features_ and there are many types of simple features, including multipolygons, lines, and points. If `geom_sf()` encounters a point, it will plot a point.

```{r}
ca_counties %>% 
  ggplot() +
  geom_sf() +
  geom_sf(
    color = "blue", 
    data = . %>% mutate(geometry = st_centroid(geometry))
  )
```

You can also use `geom_sf_label()` or `geom_sf_text()` to add text to your maps.

```{r}
lowest_density <-
  ca_counties %>% 
  top_n(n = -1, wt = density)

ca_counties %>% 
  mutate(density = pmin(density, 700)) %>% 
  ggplot() +
  geom_sf() +
  geom_sf(
    color = "blue",
    data = lowest_density %>% mutate(geometry = st_centroid(geometry))
  ) +
  geom_sf_label(
    aes(label = str_remove(name, ", California")),
    hjust = -0.05,
    vjust = -0.1,
    data = lowest_density
  )
```


## Zooming

Just as you can use `coord_cartesian()` to zoom in on a normal plot, you can use `coord_sf()` to zoom in on your map.

Let's just plot Hawaii. We'll use `projection = "longlat"` because Hawaii covers only a small area. Also, setting `project = "albers"` would scale Hawaii and place it under Texas.

```{r}
hawaii <-
  ussf::boundaries(
    geograph = "cz", 
    resolution = "500k", 
    projection = "longlat"
  ) %>% 
  filter(state == "HI")

hawaii %>% 
  ggplot() +
  geom_sf()
```

We can use `coord_sf()` to zoom in on the main islands.
  
```{r}
hawaii %>% 
  ggplot() +
  geom_sf() +
  coord_sf(
    xlim = c(-160.5, -154.7), 
    ylim = c(18.9, 22.25)
  )
```
